---
title: "Notes for the ISIMIP CMIP6 data manipulations"
author: "Gerald Nelson"
date: "4/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown




# Outputs for crops
The purpose of this exercise is to provide a quantitative assessment of the impacts of climate change on key elements of food security, and in particular to identify potential limits to adaptation. Results are generated for each of the 5 ISIMIP models for four time periods - 2001-2010 (using observed data), 2021-2030, 2051-2060 and 2091-2100.

As of 8 April 2020, the following outputs are being generated. 

* Number of days in each month where 
+ the minimum temperature is below zero degrees Celsius. The R script that does this is 

+ the maximum temperature is above the temperature at which a crop would suffer yield-reducing damage for 13 crops. The R script that does this is cropExtremesCMIP6.R

+ the maximum temperature is in the optimum temperature range for a crop, for 13 crops. The R script that does this is cropExtremesCMIP6.R

* the number of chilling hours in the northern hemisphere and the southern hemisphere. The R script that does this is chilling.R

# Outputs for animals
For the following animals - cattle, sheep, goat, yak, swine, layer, and broiler - whether the monthly average maximum temperature is greater than the lethal temperature value. 
The R script that does this is THIcalcs.R

The animal calculations have two parts. First, estimate the THI value for each pixel. Second, identify those pixels whose THI values are greater than the extreme damage break point.

# Technical issues
# Input data
The climate data used in these scripts is prepared initially by the ISIMIP project [link](www.isimip.org). Although a bit of analysis is done with ISIMIP data from the CMIP5 collection, most of it is from CMIP6. We're using the ISIMIP3b output data sets [link](https://www.isimip.org/news/isimip3ab-protocol-released/). It includes data from 5 earth system models (ESMs) and three scenarios, called ssp126, ssp370 and ssp585. More information is available here [link](https://protocol.isimip.org/protocol/ISIMIP3b/agriculture.html). The data have been bias-corrected to facilitate comparisions over time. 

The data from ISIMIP are daily values for a 1/2 degree grid of the whole world. We use a land only dataset (that also excludes Antarctica) produced by Christoph Mueller of PIK for our analysis

## Change units
Two of the raw variables (in netcdf files, with suffix .nc) from ISIMIP come in units that are not condusive for food security modeling. 

* Temperature (tasmax and tasmin) has units of degrees Kelvin. These can be converted to degrees Celsius by subtracting 273.15.

* Precipitation (pr) has units of kg m−2 s-1 (kilograms of rain per square meter per second). They can be converted to mm per day by mulitiplying by 86400.

Any time a netcdf file is read in it should be immediately converted as needed. The function fixUnits(var, ncin.brick) in the globallyUsed.R file will do this for both temperature and precipitation.

## File formats used
Any climate or related dataset that is stored as a geotiff il (suffix of .tif) has already implemented these conversions.


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
